---
title: Using Pipes in R
author: Nate Byers
date: '2020-01-05'
slug: using-pipes-in-r
categories:
  - R
tags: []
---

If you've been introduced to R as a simple way to do data analysis you might have
come across this strange operator, `%>%`. It's called a pipe because it passes
data from one function to another. Here's an example of subsetting and 
transforming data using the pipe from the magrittr package:

```{r}
library(magrittr)
dat <- airquality %>%
  subset(Ozone > 40) %>%
  transform(Celsius = (Temp - 32) * (5/9)) %>%
  head()
dat
```

The first line can be read, "I'm going to make a new object called `dat` and it's
going to start with the `airquality` data frame". The `%>%` at the end of the 
first line pipes the data frame to the next line, which is the `subset` function.
If you look at the 
[documentation](http://stat.ethz.ch/R-manual/R-devel/library/base/html/subset.html) 
for `subset`, the first argument is `x`, an "object to be subsetted". The `%>%`
takes the data frame immediately before it and places it in the first argument
of the function immediately following it. So `airquality` becomes the object
to be subsetted in the `subset` function.

Since the pipe has already assigned a data frame to the first argument of `subset`,
the next argument in the function is a logical expression that is used to select
rows to keep (i.e., subset the data frame). I want to keep all rows where
the ozone values are above 40. 

Once the concept sinks in, you can easily read the rest of the code. The output
of `subset` is piped to the first argument of `transform`. The argument that I 
have inside of `transform` is assigned to the second argument, and the output
of `transform` is passed on to the first argument of `head`.

# dplyr

So why use the pipe? For one thing, you avoid reassigning the data frame every
time you change it. Here's the subset/transformation from above without the 
pipes.

```{r}
dat <- subset(airquality, Ozone > 40) 
dat <- transform(dat, Celsius = (Temp - 32) * (5/9)) 
dat <- head(dat)
dat
```

You not only avoid reassigning the data frame every time, but you don't have to
type the data frame object as the first argument in each function.

Admittedly, the amount of typing being saved is minimal. The other main reason
to use pipes is the benefit of chaining dplyr functions together. Those functions
were written with the pipe in mind. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
dat <- airquality %>%
  filter(Ozone > 40) %>%
  mutate(Celsius = (Temp - 32) * (5/9)) %>%
  head()
dat
```

Once you get used to using the pipe, you gain the ability to quickly read a 
chain of dplyr functions. And this can speed up your production significantly.